preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-06 Manuscript/manuscript.Rmd
edits

Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associated Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by TAD calling algorithms. To aid in the precise identification of TAD boundaries we developed preciseTAD, a machine learning framework that leverages the spatial relationship of many high resolution ChIP-seq defined genomic elements, coupled with density-based clustering  and scalable partitioning techniques. Our framework precisely predicts chromosome-specific TAD boundaries on multiple cell types. We show that known molecular drivers of 3D chromatin including CTCF, RAD21, and SMC3 are more enriched at our predicted TAD boundaries compared to the boundaries identified by the popular ARROWHEAD TAD caller. Our results provide useful insights into the 3D organization of the human genome.

# ABSTRACT (MD version)

**Background.** Topologically Associated Domains (TADs) represent fundamental building blocks of the three-dimensional (3D) genome organization and regulation. Several epigenetic marks, such as CTCF, are highly enriched at TAD boundaries, suggesting a strong epigenetic component of TAD boundary formation. Several methods have been developed to leverage the growing amount of (epi)genomic annotation data for predicting TAD boundaries; however, they overlooked key characteristics of the data, such as spatial associations between a genomic annotation and a boundary. Furthermore, the number of regions containing TAD boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUROC) curve, Precision-Recall curve (AUPRC), F1, Matthews Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

**Results.** Using different feature engineering and resampling strategies, coupled with elastic net-based feature selection, we systematically compared the performance of models predicting the location of TAD boundaries (Gm12878 and K562 Hi-C data at 10kb to 100kb resolutions). Precision-Recall curves were found to be best suited for evaluating model performance in class imbalance settings. Random undersampling and SMOTE-resampled data performed best for optimal parameter selection. Transcription Factor Binding Sites (TFBSs) were the best type of genomic annotations for predicting TAD boundaries when directly compared with histone modifications and chromatin states; we identified well-known CTCF, SMC3, RAD21, and several novel TFBSs as best predictors of TAD boundaries. Importantly, spatial relationships (distance between TAD boundaries and genomic annotations) were best predictors of TAD boundaries as compared with commonly used overlap counts or percent overlap measures.

**Conclusions.** Our results outline strategies for predictive modeling of 3D genomic domains using 1D genome annotation data. The accurate modeling of TAD boundaries will improve our understanding of how epigenetics shapes the 3D structure of the genome.

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). PEAKACHU chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly).

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

We use called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publicly available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        ARROWHEAD      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     ARROWHEAD         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     ARROWHEAD         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     ARROWHEAD         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     ARROWHEAD         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        PEAKACHU       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for ARROWHEAD and PEAKACHU on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions with results consistent for K562 (Supplementary Figure 2). Random forest models built on PEAKACHU-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Furthermore, 5 kb resolution genomic bins led to optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 ARROWHEAD TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using ARROWHEAD ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on PEAKACHU resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). [???need concluding sentence]

![**Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.](Figures/figure4.png)

@@

Using our optimally built random forest model trained on ARROWHEAD/PEAKACHU boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained (using the optimal characteristics outlined above) on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using ARROWHEAD and PEAKACHU ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than ARROWHEAD on both cell lines (Table 1, Supplementary Table 2). This can be attributed to ARROWHEAD's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). Likewise, ARROWHEAD often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than PEAKACHU (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting PEAKACHU which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus ARROWHEAD and PEAKACHU respectively) (Figure 5D, Supplementary Figure 7).

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, ARROWHEAD, and PEAKACHU after flanking them by the resolution of the Hi-C data. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving PEAKACHU were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for ARROWHEAD and PEAKACHU alone (Figure 5E) [???should I mention jaccard or scrap it? significantly different for arrowhead but not peakachu].

<!--Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.-->

<!--We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).-->

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.**(A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the predicted probability of each base being a TAD boundary. A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the preciseTAD boundary region (PTBR). (B) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (C) around Peakachu loop boundaries and preciseTAD for the entire genome. (D) Log 2 genomic distances from boundaries to nearest CTCF sites. (E) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure5.png)

@@

The majority of boundaries predicted by the ARROWHEAD-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (PEAKACHU) are more abundant, while TAD boundaries (ARROWHEAD) represent the narrower level of the 3D chromatin organization. Together with the results shown on Figure 4, the high overlap between boundaries predicted by ARROWHEAD- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and ChIP-seq defined genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between boundary regions and annotations, coupled with random under-sampling help to form the predictor space resulting in optimal model performance. <!--We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. The model predicts probability for each base to be a baoundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution.-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 4-5 [??? Merge them into one Excel with two sheets - same message]). Our aim was to have a balanced metric that was sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it? for MD]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

>>>>

**Background** Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associated Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by conventional TAD calling algorithms.

**Methods** To aid in the precise identification of TAD boundaries we developed preciseTAD, a random forest based framework that leverages the spatial relationship of high resolution ChIP-seq defined genomic elements, coupled with density-based clustering and scalable partitioning techniques. We benchmarked our method against a popular TAD-caller and a novel chromatin loop prediction algorithm on multiple cell lines.

**Results** We show that our framework performs well when predicting both TAD and chromatin loop boundaries. Compared to established techniques, boundaries predicted by preciseTAD are more enriched for CTCF, RAD21, SMC3, and ZNF143 sites. Likewise, preciseTAD boundaries were more conserved across cell lines, highlighting their biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using 1D genome annotation data. The precise identification of TAD boundaries will improve our understanding of how epigenetics shapes the 3D structure of the genome.

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools Arrowhead [@durand2016juicebox] and Peakachu [???change to lowercase] [@salameh2019supervised] respectively. Arrowhead [???change to lowercase] TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly).

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both Arrowhead and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

We use called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publicly available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the Arrowhead pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that Arrowhead data constitute the main results. For MD]

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions with results consistent for K562 (Supplementary Figure 2). Random forest models built on Peakachu-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Furthermore, 5 kb resolution genomic bins led to optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 Arrowhead TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). [???need concluding sentence]

![**Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.](Figures/figure4.png)

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained (using the optimal characteristics outlined above) on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7).

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E) [???should I mention jaccard or scrap it? significantly different for arrowhead but not peakachu].

<!--Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted Arrowhead and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.-->

<!--We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than Arrowhead boundaries, thus implying stronger invariance toward resolution (Figure 5E).-->

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.**(A) The location of preciseTAD (green) vs. Arrowhead (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the predicted probability of each base being a TAD boundary. A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the preciseTAD boundary region (PTBR). (B) Enriched heatmaps comparing the enrichment of CTCF signal around Arrowhead TAD boundaries and preciseTAD boundaries, and (C) around Peakachu loop boundaries and preciseTAD for the entire genome. (D) Log 2 genomic distances from boundaries to nearest CTCF sites. (E) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using Arrowhead; top left), Arrowhead (top right), preciseTAD (using Peakachu; bottom left), and Peakachu (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure5.png)

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results shown on Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and ChIP-seq defined genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between boundary regions and annotations, coupled with random under-sampling help to form the predictor space resulting in optimal model performance. <!--We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. The model predicts probability for each base to be a baoundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution.-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 4-5 [??? Merge them into one Excel with two sheets - same message]!!!will do when ready). Our aim was to have a balanced metric that was sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it? for MD]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent].  [??? So what?]. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. [??? So what?]
